The competing effects of the immune system and EMT on epithelial cancers

### 2019-04-18 Paper/Paper.tex
update discussion

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo2017multiscale}. 
The agents in the model are the tissue cells with tumorigenic potential.
These can either be mutation-free, or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
If the pathway is mutated, the component is colored red and is marked with `X'.
A single mutation is sufficient for the cell to be considered as a mutant cell.
\par
We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
The cytokine TGF-$\beta$ is continuous within the tumor microenvironment (TME), i.e. the system is well-mixed.
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
In Figure \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
MIE is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
MGA is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As the level of mesenchymal immune evasion (MIE) is increased, the Time to Cancer decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.

Second, as the rate of mesenchymal growth arrest (MGA) increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.

Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
In Figure \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal phenotype will better protect the cancer and decrease the Time to Cancer.

@@

\section{Discussion}\label{Discussion}
Despite the intense interest in cancer and the immune system, and, separately, in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been any model developed that combines all three of these components: cancer, the immune system, and EMT. We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$. We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
\par
We found that this model recapitulated cancer-free survival dynamics, and via a global (one-at-a-time) parameter sensitivity analysis, we identified parameters exerting key control over model behavior. Focussing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer. However varying the level of inflammation leads to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the Time to Cancer.
\par
To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across two groups: we found that the presence of an EMT signature increased differences in survival between groups in combination with inflammation. We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model. There are, of course, many tissue trajectories that all result in cancer, and analysis of the transient cell dynamics in tissues bound to become cancerous might tease out \tcr{which are indicators of cancer.}
\par

%%%% AM: Yes please add a para on this. 
%In addition, I could discuss the implications from the TGF-$\beta$ section.
%Any discussion on this topic, however, would need to acknowledge that regulation of TGF-$\beta$ might also have other, serious side effects, especially if the treatment is not local.
%However, that might also be true for the inflammatory results.
%In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.
These promising results lead us to believe that continuing to study and improve the model will lead to further understanding of the competing roles of the immune system and EMT on epithelial cancers.
Developing the inflammation module of our model stands out as an attractive next step.
This is because we have seen that the inflammatory state of the TME had some of the most significant effects on Time to Cancer (see Figures \ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}BD).
The cycling nature of the scheme we chose is meant to synthesize the disparate inputs that affect the local tissue and output a simple and interpretable result.
However, many of the agents explicitly considered in the model actively contribute to the inflammatory state of the TME, providing a basis for connecting these two pieces of our model so that, for example, as mutant cells start growing, inflammation begins to rise.
In this way, a more robust inflammatory state could be modeled that would allow for more informative analysis.
\par
Coupling nicely with this improvement, the natural anti-inflammatory role of Tregs would give another layer to the competing roles of the immune system.
One consequence of our current model is that it appears that fewer Tregs means longer Time to Cancer, implying that simply eliminating Tregs is best.
Adding in this main effector function of Tregs into the model might help change this oversimplified analysis and add depth to the understanding of Tregs we glean from the model.
%Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
%One intriguing aspect could be the anti-inflammatory function of Tregs as they seek to restore homeostasis.
\par
In addition, the role that TGF-$\beta$ plays throughout the TME sets it up as a prime target of research.
We saw in Figure \ref{fig:FirstSurvivalCurves}E that below some threshold, reduction of TGF-$\beta$ increases Time to Cancer.
Thus, our model suggests that reducing quantities of TGF-$\beta$ in the TME will benefit patient survival.
However, it must be noted that TGF-$\beta$ is a powerful cytokine implicated in many cellular processes and attempts to shift the concentrations of TGF-$\beta$ in the TME or throughout the body could have undesirable side effects.
With an aggressive disease like cancer, though, more aggressive therapies are warranted.
\par
%%% AM: please expand on this. how do we give it a larger role? What does "more robust" mean? how to further explore the anti-inflamm function of the Tregs? 
\par
Finally, Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system. It is our view that the role of decanalization \cite{gibson09_decanalization} in during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}. Progress in cancer therapies has proven to be attainable even despite great challenges, for which the complexity of the disease is often in large part responsible. As we move towards a new generation of immunotherapies, it is these very complexities that we must better understand in order to eradicate or control the disease.

>>>>

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo2017multiscale}. 
The agents in the model are the tissue cells with tumorigenic potential.
These can either be mutation-free, or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
If the pathway is mutated, the component is colored red and is marked with `X'.
A single mutation is sufficient for the cell to be considered as a mutant cell.
\par
We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
The cytokine TGF-$\beta$ is continuous within the tumor microenvironment (TME), i.e. the system is well-mixed.
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
MIE is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
MGA is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As the level of mesenchymal immune evasion (MIE) is increased, the Time to Cancer decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.

Second, as the rate of mesenchymal growth arrest (MGA) increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig. \ref{fig:FirstSurvivalCurves}B).
This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.

Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
In Fig. \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal phenotype will better protect the cancer and decrease the Time to Cancer.

@@

\section{Discussion}\label{Discussion}
Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these components: cancer, the immune system, and EMT. We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$. We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
\par
We found that this model recapitulated cancer-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior. Focusing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer. However varying the level of inflammation leads to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the Time to Cancer.
\par
To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across two groups: we found that the presence of an EMT signature increased differences in survival between groups in combination with inflammation. We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model. There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in pre-cancerous tissues is a pressing need to shed insight into cellular biomarkers of cancer.
\par

%%%% AM: Yes please add a para on this. 
%In addition, I could discuss the implications from the TGF-$\beta$ section.
%Any discussion on this topic, however, would need to acknowledge that regulation of TGF-$\beta$ might also have other, serious side effects, especially if the treatment is not local.
%However, that might also be true for the inflammatory results.
%In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.

These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done. Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs. \ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}B-D).
The cycling nature of the scheme we chose is meant to synthesize the disparate inputs that affect the local tissue and output a simple and interpretable result.
However, many of the agents explicitly considered in the model actively contribute to the inflammatory state of the TME, providing a basis for connecting these two pieces of our model so that, for example, as mutant cells start growing, inflammation begins to rise.
In this way, a more robust inflammatory state could be modeled that would allow for more informative analysis.
\par
Coupling nicely with this improvement, the natural anti-inflammatory role of Tregs would give another layer to the competing roles of the immune system.
One consequence of our current model is that it appears that fewer Tregs means longer Time to Cancer, implying that simply eliminating Tregs is best.
Adding in this main effector function of Tregs into the model might help change this oversimplified analysis and add depth to the understanding of Tregs we glean from the model.
%Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
%One intriguing aspect could be the anti-inflammatory function of Tregs as they seek to restore homeostasis.
\par
In addition, the role that TGF-$\beta$ plays throughout the TME sets it up as a prime target of research.
We saw in Fig. \ref{fig:FirstSurvivalCurves}E that below some threshold, reduction of TGF-$\beta$ increases Time to Cancer.
Thus, our model suggests that reducing quantities of TGF-$\beta$ in the TME will benefit patient survival.
However, it must be noted that TGF-$\beta$ is a powerful cytokine implicated in many cellular processes and attempts to shift the concentrations of TGF-$\beta$ in the TME or throughout the body could have undesirable side effects.
With an aggressive disease like cancer, though, more aggressive therapies are warranted.
\par
%%% AM: please expand on this. how do we give it a larger role? What does "more robust" mean? how to further explore the anti-inflamm function of the Tregs? 
\par
Finally, Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system. It is our view that the role of decanalization \cite{gibson09_decanalization} in during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}. Progress in cancer therapies has proven to be attainable even despite great challenges, for which the complexity of the disease is often in large part responsible. As we move towards a new generation of immunotherapies, it is these very complexities that we must better understand in order to eradicate or control the disease.
